\section{Meaning as Pointers into a Buffer}\label{sec:pointers}

Now that we know the serialisation format, we can give a meaning
to constructor and data descriptions as pointers into a buffer.
%
For reasons that will become apparent in \Cref{sec:bufferfold}
when we start programming over serialised data in a correct-by-construction
manner, our types of `pointers' will be parameterised not only
by the description of the type of the data stored but also by a
runtime-irrelevant inductive value of that type.
%
For now, it is enough to think of these indices as a lightweight
version of the `points to' assertions used in separation
logic~\citep{DBLP:conf/lics/Reynolds02}
when reasoning about imperative programs.
%
We expand on this analogy in \Cref{appendix:hoare} where we
also discuss the connection with the combinators defined
in \Cref{sec:poking}.

\subsection{Tracking Buffer Positions}

We start with the definition of the counterpart to \IdrisType{Mu}
for serialised values.

\ExecuteMetaData[SaferIndexed.idr.tex]{pointermu}

A tree sitting in a buffer is represented
by a record packing the buffer, the position at which the tree's
root node is stored, and the size of the tree.
%
Note that according to our serialisation format the size is not stored
in the file but using the size of the buffer, the stored offsets,
and the size of the static data we will always
be able to compute a value corresponding to it.

\ExecuteMetaData[SaferIndexed.idr.tex]{elem}

The counterpart to a \IdrisFunction{Meaning} stores additional information.
%
For a description of type (\IdrisType{Desc} \IdrisBound{r} \IdrisBound{s} \IdrisBound{o})
on top of the buffer, the position at which the root of the meaning resides,
and the size of the layer we additionally have a vector of \IdrisBound{o} offsets
that allow us to efficiently access any value we want.

\subsection{Writing a Tree to a File}\label{sec:writetofile}

Once we have a pointer to a tree in a buffer, we can easily write it to a
file be it for safekeeping or sending over the network.

\ExecuteMetaData[SaferIndexed.idr.tex]{writeToFile}

We first start by reading the size of the header stored in the buffer.
%
This allows us to compute both the \IdrisBound{start} of the data block
as well as the size of the buffer (\IdrisBound{bufSize}) that will
contain the header followed by the tree we want to write to a file.
%
We then check whether the position of the pointer is exactly the beginning
of the data block.
%
If it is then we are pointing to the whole tree and the current buffer can
be written to a file as is.
%
Otherwise we are pointing to a subtree and need to separate it from its
surrounding context first.
%
To do so we allocate a new buffer of the right size and use the
standard library's \IdrisFunction{copyData} primitive to copy the raw bytes
corresponding to the header first, and the tree of interest second.
%
We can then write the buffer we have picked to a file and happily succeed.



Now that we have pointers and can save the tree they are standing for,
we are only missing the ability to look at the content they are pointing to.
%
But first we need to introduce some basic tools
to be able to talk precisely about this stored content.
