
\subsection{The Singleton type}\label{sec:datasingleton}

The \IdrisType{Singleton} family has a single constructor
which takes an argument \IdrisBound{x} of type \IdrisBound{a},
its return type is indexed precisely by this \IdrisBound{x}.

\ExecuteMetaData[Data/Singleton.idr.tex]{singleton}

More concretely this means that a value of type
(\IdrisType{Singleton} $t$) has to be a runtime relevant
copy of the term $t$.
%
Note that \idris{} performs an optimisation similar to Haskell's
\texttt{newtype} unwrapping: every data type that has a single
non-recursive constructor with only one non-erased argument
is unwrapped during compilation.
%
This means that at runtime the
\IdrisType{Singleton} / \IdrisData{MkSingleton} indirections
will have disappeared.

We can define some convenient combinators to manipulate
singletons.
%
We reuse the naming conventions typical of applicative
functors which will allow us to rely on \idris{}'s automatic
desugaring of \emph{idiom brackets}~\citep{DBLP:journals/jfp/McbrideP08}
into expressions using these combinators.

First \IdrisFunction{pure} is a simple alias for \IdrisData{MkSingleton},
it turns a runtime-relevant value \IdrisBound{x} into a singleton for
this value.

\ExecuteMetaData[Data/Singleton.idr.tex]{pure}

Next, we can `map' a function under a \IdrisType{Singleton} layer: given
a pure function \IdrisBound{f} and a runtime copy of \IdrisBound{t} we
can get a runtime copy of (\IdrisBound{f} \IdrisBound{t}).

\ExecuteMetaData[Data/Singleton.idr.tex]{fmap}

Finally, we can apply a runtime copy of a function \IdrisBound{f}
to a runtime copy of an argument \IdrisBound{t}
to get a runtime copy of the result (\IdrisBound{f} \IdrisBound{t}).

\ExecuteMetaData[Data/Singleton.idr.tex]{ap}

As we mentioned earlier, \idris{} automatically desugars idiom brackets
using these combinators.

\begin{remark}[Idiom Brackets as Sugar]
  Idiom brackets let us use the standard whitespace-based application
  typical of pure functions to performs computations in an \emph{Applicative}
  contexts.

  In other words, during elaboration an atomic expression
  \IdrisKeyword{[|} \IdrisBound{x} \IdrisKeyword{|]}
  will be desugared to (\IdrisFunction{pure} \IdrisBound{x})
  while a compound expression
  \IdrisKeyword{[|} \IdrisBound{f} \IdrisBound{t1} $\cdots$ \IdrisBound{tn} \IdrisKeyword{|]}
  will become (\IdrisBound{f} \IdrisFunction{<\$>} \IdrisBound{t1} \IdrisFunction{<*>} $\cdots$ \IdrisFunction{<*>} \IdrisBound{tn}).
\end{remark}

This built-in handling of idiom brackets lets us apply
\IdrisType{Singleton}-wrapped values almost as seamlessly
as pure values.
