
\section{Interlude: Views and Singletons}\label{sec:view}

The precise indexing of pointers by a runtime-irrelevant copy of the value
they are pointing to means that inspecting the buffer's content should
not only return runtime information but also refine the index to reflect
that information at the type-level.
%
As a consequence, the functions we are going to define in the following
subsections are views.

\subsection{Views}

A view
in the sense of Wadler~\citep{DBLP:conf/popl/Wadler87},
and subsequently refined by McBride and McKinna~\citep{DBLP:journals/jfp/McBrideM04}
for a type $T$ is a type family $V$ indexed by $T$ together
with a function which maps values $t$ of type $T$ to values of type
$V\,t$.
%
By inspecting the $V\,t$ values we can learn something about the
$t$ input.
%
The prototypical example is perhaps the `snoc` (`cons' backwards) view
of right-nested lists as if they were left-nested.
We present the \IdrisType{Snoc} family below.

\ExecuteMetaData[Snoc.idr.tex]{Snoc}

By matching on a value of type
(\IdrisType{Snoc} \IdrisBound{xs}) we get to learn
either that \IdrisBound{xs} is empty (\IdrisData{Lin}, nil backwards)
or that it has an initial segment \IdrisBound{init} and a last element
\IdrisBound{last} (\IdrisBound{init} \IdrisData{:<} \IdrisBound{last}).
%
The function \IdrisFunction{unsnoc} demonstrates that we can always
\emph{view} a \IdrisType{List} in a \IdrisType{Snoc}-manner.

\ExecuteMetaData[Snoc.idr.tex]{unsnoc}

\begin{remark}[With-Abstraction]
  The \IdrisKeyword{with} construct allows programmers to locally
  define an anonymous auxiliary function taking an extra argument
  compared to its parent.
  %
  By writing (\IdrisKeyword{with} \IdrisKeyword{(}$e$\IdrisKeyword{)})
  we introduce such an auxiliary function an immediately apply it to $e$.
  %
  The nested clauses that immediately follow each take an extra pattern
  which matches over the possible values of $e$.
  %
  If the left-hand side of the auxiliary function is the same as that of
  its parents bar the pattern for the newly added argument, we can use the
  elision notation (\IdrisKeyword{\_} \IdrisKeyword{|}) to avoid having to
  repeat ourselves.

  In other words the following definition of \IdrisFunction{f} using a
  \IdrisKeyword{with} construct with the elision notation

  \ExecuteMetaData[With.idr.tex]{fsugar}

  \noindent is equivalent to the following desugared version where
  the auxiliary function taking an extra argument has been lifted
  to the toplevel.

  \ExecuteMetaData[With.idr.tex]{fdesugar}

\end{remark}

In the above code we performed a recursive call on
(\IdrisFunction{unsnoc} \IdrisBound{xs}) and distinguished
two cases: when the view returns the empty snoclist \IdrisData{[<]}
and when it returns an \IdrisBound{init}ial segment together with the
\IdrisBound{last} element.
%
Because we are using a view, matching on these constructors actually
refined the shape of the parent clause's argument \IdrisBound{xs}.
We do not need to spell out its exact shape in each branch because
we were careful to only introduce \IdrisBound{xs} as a name alias
using an as-pattern while letting the actual pattern be a catch-all
pattern (\IdrisBound{xs}\IdrisKeyword{@}\IdrisKeyword{\_}).
%
This is a common trick to make working with views as lightweight as
possible.

Here we defined \IdrisType{Snoc} as an inductive family but it can
sometimes be convenient to define the family recursively instead,
in which case the \IdrisType{Singleton} inductive family can
help us connect runtime values to their
runtime-irrelevant type-level counterparts.

\input{singleton}

We are now equipped with the appropriate notions and definitions to
look at a buffer's content.
