\section{Our Universe of Descriptions}\label{sec:desc}

We first need to pin down the domain of our discourse.
%
To talk generically about an entire class of datatypes
without needing to modify the host language
we have decided to perform a universe
construction~\citep{DBLP:journals/njc/BenkeDJ03, DBLP:phd/ethos/Morris07, DBLP:conf/icfp/LohM11}.
%
That is to say that we are going to introduce an inductive type
defining a set of codes together
with an interpretation of these codes as bona fide
host-language types.
%
We will then be able to program generically over the universe of
datatypes by performing induction on the type of
codes~\citep{DBLP:conf/tphol/PfeiferR99}.

The universe we define is in the tradition of
a sums-of-products vision of inductive types~\citep{DBLP:conf/popl/JanssonJ97}.
%
In our setting, constructors are essentially arbitrarily nested tuples of
values of type unit,
bytes,
and recursive substructures.
%
A datatype is given by listing a choice of constructors.

\subsection{Interlude: Serialisation Formats}\label{sec:interludeserialisation}

Before giving a precise definition of the descriptions, let us think
a little bit about the features we would like to see in our serialisation
format and see how this informs our design.
%
We reproduce below the naïve encoding of a binary tree with bytes stored
in its nodes we gave in the introduction. It has both strengths and
weaknesses.

\begin{center}
  \usestt
      \texttt{01 $\overbrace{\texttt{01 $\underbrace{\texttt{01 00 \IdrisData{01} 00}}_{\texttt{(node leaf \IdrisData{1} leaf)}}$
    \IdrisData{05} 00}}^{\texttt{(node (node leaf \IdrisData{1} leaf) \IdrisData{5} leaf)}}$
    \IdrisData{0a} 01 00 \IdrisData{14} 00}
\end{center}

One of its strength is the flattened encoding of nodes:
the left subtree, byte stored, and right subtree are all stored
contiguously after the tag \texttt{01} announcing that a \texttt{node} is present.
%
This approach can be generalised to any type with a unique constructor:
there is no need to store a byte corresponding to the constructor for the
unit type, or for the constructor to a pair type. These unambiguous
constructors can be reinserted during the type-directed decoding phase.
%
In practice this means that our encoding will not retain the nesting
structure described by a type. Concretely, $((a, b), c)$ and $(a, (b, c))$
will have the same representations: the encoding of $a$, followed by the
encoding of $b$ and last that of $c$.
%
Similarly, values of type $()$ will be eluded entirely: values of type
$((), a, ())$ and $a$ will have the same serialised representation, the
encoding of the value of type $a$.


The major weakness of the naïve format we presented above is the
inability to process a node's right subtree without having first
processed its left subtree.
%
This is due to the fact that the subtree's size is not statically
known and that the serialisation format does not store an offset
for it.
%
Consequently, if we wanted to fetch the rightmost byte stored
in the tree, we would have to traverse the entire encoding. That
represents an exponential slow-down compared to what we could achieve
with a different encoding allowing for random access.

Based on these observations, we envision a serialisation format
that will permit random access to all of the arguments of a node.
Such a format will have to store an offset for all of a node's
arguments whose size is not statically known and which we may want
to jump past i.e. all of them except a constructor's very last argument.

\subsection{Descriptions}

This leads us to the inductive family \IdrisType{Desc} declared below
as the type of constructor descriptions.

\ExecuteMetaData[Serialised/Desc.idr.tex]{desctype}

\begin{remark}[Separate Declaration and Definition]
  \idris{} lets us declare a datatype or function first and only
  define it later. This is convenient for blocks of mutually-defined
  types and functions but can also be used, like here, to focus on
  the type signature first.
\end{remark}

This family has three indices corresponding to three crucial
invariants being tracked.
%
First, an index telling us whether the current description
is being used in the \IdrisBound{rightmost} branch of the overall
constructor description.
%
Second, the \IdrisBound{static}ally known size of the described data
in the number of bytes it occupies.
%
Third, the number of \IdrisBound{offsets} that need to be stored to
compensate for subterms not having a statically known size.
%
The reader should think of \IdrisBound{rightmost} as an `input' index
(the context in which the family is used tells it whether it is in
a rightmost branch)
whereas \IdrisBound{static} and \IdrisBound{offsets} are `output' indices
(the family's own constructors each determine their respective sizes).

Next we define the family proper by giving its four constructors.

\ExecuteMetaData[Serialised/Desc.idr.tex]{desc}

Each constructor can be used anywhere in a description so their index
tracking whether we are in the rightmost branch can be an arbitrary
boolean \IdrisBound{r}.

\IdrisData{None} is the description of values of type unit. The static
size of these values is zero as no data is stored in a value of type unit.
Similarly, they do not require an offset to be stored as we statically
know their size.

\IdrisData{Byte} is the description of bytes.
%
Their static size is precisely one byte, and they do not require an
offset to be stored either.

\IdrisData{Prod} gives us the ability to pair two descriptions together.
As explained earlier pairs have a unique constructor and so their encoding
will not consume any additional bytes.
Consequently their static size and number of offsets are the respective
sums of the static sizes and numbers of offsets of each subdescription.
%
The description of the left element of the pair will never be in the
rightmost branch of the overall constructors description and so its
index is \IdrisData{False} while the description of the right element
of the pair is in the rightmost branch precisely whenever the whole pair
is; hence the propagation of the \IdrisBound{r} arbitrary value from the
return index into the description of the right component.

Last but not least, \IdrisData{Rec} is a position for a subtree.
We cannot know its size in bytes statically and so we decide to store
an offset unless we are in the rightmost branch of the overall description.
%
Indeed, there are no additional constructor arguments behind the rightmost
one and so we have no reason to skip past the subterm. Consequently we
do not bother recording an offset for it.


\subsection{Constructors}

We represent a constructor as a record packing together
a name for the constructor,
the description of its arguments (which is, by virtue of
being used at the toplevel, in rightmost position),
and the values of the \IdrisFunction{static} and
\IdrisFunction{offsets} invariants.

\ExecuteMetaData[Serialised/Desc.idr.tex]{constructor}

\begin{remark}[Implicit Record Fields]
  Record fields whose type declaration is surrounded by
  curly braces are implicit: they do not need to be explicitly
  mentioned when constructing the record, or when pattern-matching
  on it.
\end{remark}

Here the two invariants \IdrisFunction{static} and
\IdrisFunction{offsets} are stored as implicit fields
because as `output' indices of the \IdrisType{Desc} family
their value is easily reconstructed using
unification~\citep{DBLP:conf/tlca/AbelP11}.
%
Note that we used \IdrisData{(::)} as the name of the
constructor for records of type \IdrisType{Constructor}.
This allows us to define constructors by forming an
expression reminiscent of Haskell's type declarations:
\IdrisBound{name} \IdrisData{::} \IdrisBound{type}.
%
Returning to our running example, this gives us the following encodings for
leaves that do not store anything
and nodes that contain a left branch, a byte, and a right branch.

\noindent
\begin{minipage}[t]{.38\textwidth}
  \ExecuteMetaData[Serialised/Desc.idr.tex]{treeleaf}
\end{minipage}\hfill
\begin{minipage}[t]{.58\textwidth}
  \ExecuteMetaData[Serialised/Desc.idr.tex]{treenode}
\end{minipage}

\subsection{Datatypes}

A datatype description \IdrisType{Data}
is given by a vector (also known as
a length-indexed list) named \IdrisFunction{constructors}
and containing constructor descriptions.
%
For convenience, we store the number \IdrisFunction{consNumber}
of constructors separately so that it does not need to be
recomputed every time it is needed.
%
We additionally insist that we have a proof \IdrisFunction{fitsInBits8}
that the datatype has less than 255 constructors which will allow us
to safely store the corresponding tag in a single byte.
This is enforced using \IdrisType{So}, a type
family ensuring the boolean check it is indexed by has succeeded.

\ExecuteMetaData[Serialised/Desc.idr.tex]{data}\label{sec:datadescriptions}

\begin{remark}[Auto Implicit Record Fields]
  Implicit record fields whose type declaration is preceded by the
  \IdrisKeyword{auto} keyword are auto implicit: just like implicit
  fields they do not need to be explicitly mentioned when constructing
  the record, or when pattern-matching on it.
  Additionally, these values are not constructed by unification but
  via a builtin type-directed proof search.
\end{remark}

We can then encode our running example as a simple \IdrisType{Data}
declaration: a binary tree whose node stores bytes is described by the choice
of either a \IdrisFunction{Leaf} or \IdrisFunction{Node}, as defined above.
The proof \IdrisFunction{fitsInBits8} is constructed automatically:
the boolean test (\IdrisData{2} \IdrisFunction{<=} \IdrisData{255})
has trivially computed to \IdrisData{True}.

\ExecuteMetaData[Serialised/Desc.idr.tex]{treedesc}\label{fig:treedesc}

Now that we have a language that allows us to give a description of our
inductive types, we are going to give these descriptions a meaning as trees.
