\section{Serialised Representation}\label{sec:hexdump}

Before we can give a meaning to descriptions as pointers into a
buffer we need to decide on a serialisation format.
%
The format we have opted for is split in two parts: a header, followed
by the actual representation of the tree as a contiguous block of bytes.
%
The header contains an encoding of the description of the type of the
tree stored in the rest of the buffer.
%
It can be used when loading a file to check that a user's claim about
the presence of a serialised tree of a given type is correct.


For instance, the following binary snippet is a hex dump of a file
containing the serialised representation of a binary tree belonging to
the type we have been using as our running example.
%
The raw data is semantically highlighted:
8-bytes-long\footnote{This implicitly means that we assume
that our terms occupy less than $2^{64}$ bytes,
a very reasonable assumption in our opinion.}
\hexaoffset{offsets} in little endian format,
a \hexadesc{type} description of the stored data,
some \hexacons{nodes} of the tree
and the \hexadata{data} stored in the nodes.

\begin{center}
\input{data-hexdump}
\end{center}

More specifically, this block is the encoding of the \IdrisFunction{example}
given in the previous section and,
%
accepting that the tree's representation starts at byte \texttt{FF}, and
%
knowing that a \IdrisFunction{leaf} is represented here by \hexacons{00}
and a \IdrisFunction{node} is represented by \hexacons{01}
%
readers can check
(ignoring the offsets for now)
that the data is stored in a depth-first, left-to-right traversal of the tree
i.e. we get exactly the bit pattern we saw in the naïve encoding
presented in \Cref{sec:intro}:

\begin{center}
  \usestt
      \texttt{\IdrisFunction{01} \hexaoffset{$\cdots$} $\overbrace{\texttt{\IdrisFunction{01} \hexaoffset{$\cdots$}  $\underbrace{\texttt{\IdrisFunction{01} \hexaoffset{$\cdots$}  \IdrisFunction{00} \IdrisData{01} \IdrisFunction{00}}}_{\texttt{(\IdrisFunction{\footnotesize{node}} \IdrisFunction{\footnotesize{leaf}} \IdrisData{1} \IdrisFunction{\footnotesize{leaf}})}}$
    \IdrisData{05} \IdrisFunction{00}}}^{\texttt{(\IdrisFunction{\footnotesize{node}} (\IdrisFunction{\footnotesize{node}} \IdrisFunction{\footnotesize{leaf}} \IdrisData{1} \IdrisFunction{\footnotesize{leaf}}) \IdrisData{5} \IdrisFunction{\footnotesize{leaf}})}}$
    \IdrisData{0a} \IdrisFunction{01} \hexaoffset{$\cdots$}  \IdrisFunction{00} \IdrisData{14} \IdrisFunction{00}}
\end{center}

Let us now look at the format more closely.

\subsection{Header}

In our example, the header is as follows:
\begin{hexdump}
\hexaoffset{07 00 00 00 00 00 00 00} \hexadesc{02 00 02 03 02 01 03}
\end{hexdump}

The header consists of an 8-bytes long offset stored in little endian
followed by a binary representation of the
\IdrisType{Data} description of the value stored in the buffer.
The offset allows us to jump past the description in case we do
not care to inspect it.

This description can be useful in a big project where different
components produce and consume such serialised values:
if we change the format in one place but forget to update
it in another, we want the program to gracefully
fail to load the file using an unexpected format.
%
We detail in \Cref{sec:limitation-robust} how dependent
type providers~\citep{DBLP:conf/icfp/Christiansen13}
can help structure a software project
to prevent such issues at compile time.

The encoding of a data description starts with a byte giving us the number
of constructors\footnote{
Remember that we enforce in the definition of the \IdrisType{Data} record
type in \Cref{sec:datadescriptions}
that our descriptions cannot have more than 255 constructors. So using
a single byte here is safe.},
followed by these constructors' respective descriptions
serialised one after the other.
%
\IdrisData{None} is represented by \hexadesc{00},
\IdrisData{Byte} is represented by \hexadesc{01},
(\IdrisData{Prod} \IdrisBound{d} \IdrisBound{e}) is represented by
\hexadesc{02} followed by the representation of \IdrisBound{d} and then that of \IdrisBound{e},
and \IdrisData{Rec} is represented by \hexadesc{03}.


Looking once more at the header in the running example,
the \IdrisType{Data} description is indeed 7 bytes long like the offset states.
The \IdrisType{Data} description starts with \hexadesc{02}
meaning that the type has two constructors.
The first one is \hexadesc{00} i.e. \IdrisData{None}
(this is the encoding of the type of \IdrisFunction{Leaf}),
and the second one is \hexadesc{02 03 02 01 03} i.e.
\IdrisKeyword{(}\IdrisData{Prod} \IdrisData{Rec}
\IdrisKeyword{(}\IdrisData{Prod} \IdrisData{Byte} \IdrisData{Rec}\IdrisKeyword{))}
(that is to say the encoding of the type of \IdrisFunction{Node}).
%
According to the header, this file does contain a \IdrisFunction{Tree}.

\subsection{Tree Serialisation}\label{sec:tree-serialisation}

The design of this format was guided by a few simple principles:
\begin{enumerate}
  \item the format should be contiguous (no pointer indirections)
  \item values of type unit should occupy no space
  \item the nesting of pairs in a description should have no impact on the layout
  \item the format should support the direct access to any of a node's subtrees
\end{enumerate}
%
This last criterion will allow us to, for instance,
implement a function looking up the value stored
in the rightmost node in our running example type
of binary trees in time linear in the depth of the tree
rather than exponential if we had to traverse every subtree in order.
%
To this end each node needs to store an offset measuring the size of the
subtrees that are to the left of any relevant information.

If a given tag is associated to a description of type
(\IdrisType{Desc} \IdrisData{True} \IdrisBound{s} \IdrisBound{o})
then the representation in memory of the associated node will look something
like the following.
%
On the first line we have a description of the data layout and on the
second line we have the offset of various positions in the block with
respect to the tag's address.

\label{fig:data-layout}
\begin{center}
\input{data-layout}
\end{center}


For the data layout,
we start with the tag
then we have $o$ offsets,
and finally we have a block contiguously storing an interleaving of
subtrees and $s$ bytes
dictated by the description.
%
In this example the rightmost value in the description is a subtree and
so even though we have $o$ offsets, we actually have $(o+1)$ subtrees stored.

The offsets of the tag with respect to its own address is $0$.
The tag occupies one byte and so the offset of the block of offsets is $1$.
Each offset occupies 8 bytes and so the constructor's arguments
are stored at offset $(1+8\times{}o)$.
Finally each value's offset can be computed by adding up
the offset of the start of the block of constructor arguments,
the offsets corresponding to all of the subtrees that come before it,
and the number of bytes stored before it;
in the case of the last byte that gives $1+8\times{}o + \Sigma_{i=1}^{o}o_i + s-1$
hence the formula included in the diagram.

Going back to our running example, this translates to the following
respective data layouts and offsets for a leaf and a node.

\begin{center}
  \input{data-layout-nodes}
\end{center}


We now have a good understanding of the serialisation format we
are going to use to represent our inductive trees.
%
The next step is to define what it means to have a pointer to a
tree residing in a buffer.
