
\section{An Efficient Invariant-Rich Representation}\label{sec:efficient}

We can combine both approaches highlighted in \cref{sec:thinningsintypos}
by defining a record parameterised by a source
(\IdrisBound{sx}) and target (\IdrisBound{sy}) scopes corresponding to the two
ends of the thinnings, just like we would for the inductive family. This record
packs two numbers and a runtime irrelevant proof.

Firstly, we have a natural number called \IdrisFunction{bigEnd} corresponding
to the size of the big end of the thinning (\IdrisBound{sy}).
%
We are happy to use a (unary) natural number here because we know that \idris{}
will compile it to an unbounded integer.

Secondly, we have an integer called \IdrisFunction{encoding} corresponding to
the thinning represented as a bit vector stating, for each variable, whether
it is kept or dropped. We only care about the integer's \IdrisFunction{bigEnd}
first bits and assume the rest is set to 0.

Thirdly, we have a runtime irrelevant proof \IdrisFunction{thinning} that
\IdrisFunction{encoding} is indeed a valid encoding of size \IdrisFunction{bigEnd}
of a thinning from \IdrisBound{sx} to \IdrisBound{sy}. We will explore the
definition of the relation \IdrisType{Thinning} later on
in \cref{sec:thininginternal}.

\ExecuteMetaData[Thin.idr.tex]{thin}

The first sign that this definition is adequate is our ability to construct
any valid thinning. We demonstrate it is the case by introducing functions
that act as smart constructor analogues for the inductive family's data
constructors.

\subsection{Smart Constructors for \IdrisType{Th}}

The first and simplest one is \IdrisFunction{done}, a function that packs a pair of
\IdrisData{0} (the size of the big end, and the empty encoding) together with a proof
that it is an adequate encoding of the thinning from the empty scope to itself.
In this instance, the proof is simply the \IdrisData{Done} constructor.

\ExecuteMetaData[Thin.idr.tex]{done}

Both \IdrisFunction{keep} and \IdrisFunction{drop} are going to need to extend the
encoding with an additional bit.
%
For this purpose we introduce the \IdrisFunction{cons} function which takes a bit
$b$ and an existing encoding $bs$ and returns the new encoding $\bitcons{bs}{b}$.

\ExecuteMetaData[Data/Bits/Integer.idr.tex]{cons}

No matter what the value of the newt bit is, we start by shifting the encoding to
the left to make space for it; this gives us $\bitcons{bs}{0}$.
%
If the bit is \IdrisData{True} then we need to additionally set the bit at position
$0$ to obtain $\bitcons{bs}{1}$. Otherwise if the bit is \IdrisData{False}, we can readily
return the $\bitcons{bs}{0}$ encoding obtained by left shifting.
%
The correctness of this function is backed by two lemma:
testing the bit at index $0$ after consing amounts to returning the cons'd bit,
and shifting the cons'd encoding to the right takes us back to the unextended
encoding.

\ExecuteMetaData[Data/Bits/Integer.idr.tex]{testBit0Cons}
\ExecuteMetaData[Data/Bits/Integer.idr.tex]{consShiftR}

The \IdrisFunction{keep} smart constructor demonstrates that from a thinning from
\IdrisBound{sx} to \IdrisBound{sy} and a runtime irrelevant variable \IdrisBound{x}
we can compute a thinning from the extended source scope
(\IdrisBound{sx} \IdrisData{:<} \IdrisBound{x}) to the target scope
(\IdrisBound{sy} \IdrisData{:<} \IdrisBound{x}) where \IdrisBound{x} was kept.

\ExecuteMetaData[Thin.idr.tex]{keep}

The outer scope has grown by one variable and so we increment \IdrisFunction{bigEnd}.
%
The encoding is obtained by \IdrisFunction{cons}-ing the boolean \IdrisData{True} to
record the fact that this new variable is kept.
%
Finally, we use the two lemmas shown above to convince \idris{} the invariant
has been maintained.

Similarly the \IdrisFunction{drop} function demonstrates that we can compute a
thinning getting rid of the variable \IdrisBound{x} freshly added to the target
scope.

\ExecuteMetaData[Thin.idr.tex]{drop}

We once again increment the \IdrisFunction{bigEnd}, use \IdrisFunction{cons} to
record that the variable is being discarded and use the lemmas ensuring its
correctness to convince \idris{} the invariant is maintained.

We can already deploy these smart constructors to implement functions producing
thinnings. We use \IdrisFunction{which} as our example. It is a filter-like
function that returns the elements that satisfy a boolean predicate together with
a proof that there is a thinning embedding them back into the input snoclist.
%
If the input snoclist is empty then the output shall also be, and
\IdrisFunction{done} builds a thinning from \IdrisData{[<]} to itself.
%
If it is not empty we can perform a recursive call on the tail of the snoclist
and then depending on whether the predicates holds true of the head we can either
\IdrisFunction{keep} or \IdrisFunction{drop} it.

\ExecuteMetaData[Thin.idr.tex]{which}

We are now equipped with these smart constructors that allow us to seamlessly
build thinnings.
%
To recover the full expressive power of the inductive family, we also need to
be able to take these thinnings apart. We are now going to tackle this issue.

\subsection{Pattern Matching on \IdrisType{Th}}

The \IdrisType{View} family is a sum type indexed by a thinning. It has one
data constructor associated to each smart constructor and storing its arguments.

\ExecuteMetaData[Thin.idr.tex]{view}

The accompanying \IdrisFunction{view} function witnesses the fact that any
thinning arises as one of these three cases.

\ExecuteMetaData[Thin.idr.tex]{viewtotal}

We show the implementation of \IdrisFunction{view} in its entirety but leave
out the slightly technical auxiliary lemma it invokes.
%
The interested reader can find them in the accompanying material.
%
We will however inspect the code \IdrisFunction{view} compiles to after erasure
in \cref{sec:compiledview} to confirm that these auxiliary definitions do not
incur any additioanl runtime cost.

We first start by pattern matching on the \IdrisFunction{bigEnd} of the thinning.
If it is \IdrisData{0} then we know the thinning has to be the empty thinning.
%
Thanks to an inversion lemma called \IdrisFunction{isDone}, we can collect a lot
of equality proofs: the encoding \emph{has to} be \IdrisData{0}, the
source and target scopes \emph{have to} be the empty snoclists, and the
proof of the invariant has to be of a specific shape.
%
Rewriting by these equalities changes the goal type enough that it is ultimately
manifest that the thinning was constructed using the \IdrisFunction{done} smart
constructor and so we can use the view's \IdrisData{Done} constructor.

\ExecuteMetaData[Thin.idr.tex]{viewDone}

In case the thinning is non-empty, we need to inspect the 0-th bit of the encoding
to know whether it keeps or discards its most local variable.
%
This is done by calling the \IdrisFunction{choose} function which takes a boolean
\IdrisFunction{b} and returns a value of type
{(\IdrisType{Either} (\IdrisType{So} \IdrisBound{b}) (\IdrisType{So} (\IdrisFunction{not} \IdrisBound{b}))}
i.e. we not only inspect the boolean but also record which value we got in a proof
using the \IdrisType{So} family introduced in~\cref{sec:quantitativeTT}.

\ExecuteMetaData[Thin.idr.tex]{viewKeepDrop}

If the bit is set then we know the variable is kept.
%
And so we can invoke an inversion lemma that will once again provide us with
a lot of equalities that we immediately deploy to reshape the goal's type.
%
This ultimately lets us assemble a sub-thinning and use the view's
\IdrisData{Keep} constructor.

\ExecuteMetaData[Thin.idr.tex]{viewKeep}

If the bit is not set then we learn that the thinning was constructed using
\IdrisFunction{drop}. We can once again use an inversion lemma to rearrange
the goal and finally invoke the view's \IdrisData{Drop} constructor.

\ExecuteMetaData[Thin.idr.tex]{viewDrop}





We can readily use this function to implement pattern matching functions taking
a thinning apart. We can for instance define \IdrisFunction{kept}, the function
that counts the number of \IdrisFunction{keep} smart constructors used when
manufacturing the input thinning and returns a proof that this is exactly the
length of the source scope \IdrisBound{sx}.

\ExecuteMetaData[Thin.idr.tex]{kept}

We proceed by calling the \IdrisFunction{view} function on the input thinning
which immediately tells us that we only have three cases to consider.
%
The \IdrisData{Done} case is easily handled because the branch's refined
types inform us that both \IdrisBound{sx} and \IdrisBound{sy} are the
empty snoclist \IdrisData{[<]} whose length is evidently \IdrisData{0}.
%
In the \IdrisData{Keep} branch we learn that \IdrisBound{sx} has the shape
(\IdrisBound{\KatlaUnderscore} \IdrisData{:<} \IdrisBound{x}) and so we must return the
successor of whatever the result of the recursive call gives us.
%
Finally in the \IdrisData{Drop} case, \IdrisBound{sx} is untouched and so a
simple recursive call suffices.
%
Note that the function is correctly detected as total because the target scope
\IdrisBound{sy} is indeed getting structurally smaller at every single recursive
call.
%
It is runtime irrelevant but it can still be successfully used as a termination
measure by the compiler.

\subsection{The \IdrisType{Thinning} Relation}\label{sec:thininginternal}

We have shown the user-facing \IdrisType{Th} and have claimed that it is possible
to define smart constructors \IdrisFunction{done}, \IdrisFunction{keep},
and \IdrisFunction{drop}, as well as a \IdrisFunction{view} function.
%
This should become apparent once we show the actual definition of \IdrisType{Thinning}.

\subsubsection{Definition of \IdrisType{Thinning}}

The relation maintains the invariant between the record's
fields \IdrisFunction{bigEnd} (a \IdrisType{Nat})
and \IdrisFunction{encoding} (an \IdrisType{Integer})
and the index scopes \IdrisBound{sx} and \IdrisBound{sy}.

\ExecuteMetaData[Thin/Internal.idr.tex]{thinning}

As always, the \IdrisData{Done} constructor is the simplest.
%
It states that the thinning of size \IdrisData{Z} and encoded as the bit
pattern \IdrisData{0} is the empty thinning.

The \IdrisData{Keep} constructor guarantees that the thinning of
size (\IdrisData{S} \IdrisBound{i}) and encoding \IdrisBound{bs}
represents an injection
from (\IdrisBound{sx} \IdrisData{:<} \IdrisBound{x})
to (\IdrisBound{sy} \IdrisData{:<} \IdrisBound{x})
provided that the bit at position \IdrisData{Z} of \IdrisBound{bs}
is set, and that the rest of the bit pattern (obtained by a right shift
on \IdrisBound{bs}) is a valid thinning of size \IdrisBound{i} from
\IdrisBound{sx} to \IdrisBound{sy}.

The \IdrisData{Drop} constructor is structured the same way, except that
it insists the bit at position \IdrisData{Z} should \emph{not} be set.

We can readily use this relation to prove that some basic encoding are
valid representations of useful thinnings.

\subsubsection{Examples of \IdrisType{Thinning} proofs}

For instance, we can always define a thinning from the empty scope to
an arbitrary scope \IdrisBound{sy}.
%
The \IdrisFunction{encoding} of this thinning is \IdrisData{0} because
every variable is being discarded and its \IdrisFunction{bigEnd} is
the length of the outer scope \IdrisBound{sy}.

\ExecuteMetaData[Thin/Internal.idr.tex]{none}

The proof proceeds by induction over the outer scope \IdrisBound{sy}. If it
is empty, we can simply use the constructor for the empty thinning.
%
Otherwise we can invoke \IdrisData{Drop} on the induction hypothesis.
%
This all typechecks because (\IdrisFunction{testBit} \IdrisData{0} \IdrisData{Z})
computes to \IdrisData{False} and so the \IdrisBound{nb} proof can be constructed
automatically by \idris{}'s proof search (cf. \cref{sec:proofsearch}),
%
and (\IdrisData{0} \IdrisFunction{`shiftR`} \IdrisData{1}) evaluates to \IdrisData{0}
which means the induction hypothesis has exactly the right type.


The definition of the identity thinning is a bit more involved.
%
For a scope of size $k$, we are going to need to generate a bit pattern with
$k$ ones followed by zeros.
%
We define it in two steps.
%
First, \IdrisFunction{cofull} defines a bit pattern of $k$ zeros followed by ones
by shifting $k$ places to the left a bit pattern of ones only.
%
Then, we obtain \IdrisFunction{full} by taking the complement of \IdrisFunction{cofull}.

\noindent
\begin{minipage}{.45\textwidth}
\ExecuteMetaData[Data/Bits/Integer.idr.tex]{cofull}
\end{minipage}\hfill
\begin{minipage}{.45\textwidth}
\ExecuteMetaData[Data/Bits/Integer.idr.tex]{full}
\end{minipage}

We can then prove that the identity thinning for a scope of size \IdrisBound{n} is
represented by the pairing of (\IdrisFunction{full} \IdrisBound{n}) as the
\IdrisFunction{encoding} and \IdrisBound{n} as the \IdrisFunction{bigEnd}.

\ExecuteMetaData[Thin/Internal.idr.tex]{ones}


This proof proceeds once more by induction on the scope.
%
If the scope is empty then once again the constructor for the empty thinning will do.
%
In the non-empty case, we first appeal to an auxiliary lemma (not shown here) to
construct \IdrisBound{nb} a proof that the bit at position \IdrisData{Z} for a
non-zero \IdrisFunction{full} integer is known to be \IdrisData{True}.
%
We then need to use another lemma to cast the induction hypothesis which mentions
(\IdrisFunction{full} (\IdrisFunction{length} \IdrisBound{sx})) so that it may be
used in a position where we expect a proof talking about
(\IdrisFunction{full} (\IdrisFunction{length} (\IdrisBound{sx} \IdrisData{:<} \IdrisBound{x}))
\IdrisFunction{`shiftR`} \IdrisData{1}).

\subsubsection{Properties of the \IdrisType{Thinning} relation}

This relation has a lot of convenient properties.

First, it is proof irrelevant: any two proofs that the same
\IdrisBound{i}, \IdrisBound{bs}, \IdrisBound{sx}, and \IdrisBound{sy} are
related are provably equal.
%
Consequently, equality on \IdrisType{Th} values amounts to equality of
the \IdrisFunction{bigEnd} and \IdrisFunction{encoding} values. In particular
it is cheap to test whether a given thinning is the empty or the
identity thinning.

Second, it can be inverted~\cite{DBLP:conf/types/CornesT95} knowing only two bits:
whether the natural number is empty and what the value of the bit at position
\IdrisData{Z} of the encoding is.
%
This is what allows us to efficiently implement the \IdrisFunction{view} function
by using
these two checks and then inverting the \IdrisType{Thinning} proof to gain access
to the proof that the remainder of the thinning's encoding is valid.
%
We will see in \cref{sec:compiledview} that this leads to efficient runtime code for the view.

\subsection{Choose Your Own Abstraction Level}

Access to both the high-level \IdrisType{View} and the internal \IdrisType{Thinning}
relation means that programmers can pick the level of abstraction at which they
want to work.
%
They may need to explicitly manipulate bits to implement key operators that are
used in performance-critical paths but can also stay at the highest level for
more negligible operations, or when proving runtime irrelevant properties.

In the previous section we saw simple examples of these bit manipulations when
defining \IdrisFunction{none} (using the constant 0 bit pattern) and
\IdrisFunction{ones} using bit shifting and complement to form an initial segment
of 1s followed by 0s.

Other natural examples include the \emph{meet} and \emph{join} of two thinnings
sharing the same wider scope.
%
The join can for instance be thought of either as a function defined by induction
on the first thinning and case analysis on the second, emitting a \IdrisData{Keep}
constructor whenever either of the inputs does.
%
Or we can observe that the bit pattern in the join is exactly the disjunction of
the inputs' respective bit patterns and prove a lemma about the \IdrisType{Thinning}
relation instead.
%
This can be visualised as follows. In each column, the meet is a
$\bullet$ whenever either of the inputs is.

\[
\begin{array}{c@{~}l}
& \color{blue}{\circ}\color{orange}{\circ}\color{magenta}{\bullet}\color{teal}{\bullet}\color{lightgray}{\circ} \\
  \vee & \color{blue}{\bullet}\color{orange}{\circ}\color{magenta}{\circ}\color{teal}{\bullet}\color{lightgray}{\bullet} \\
  \hline
  & \color{blue}{\bullet}\color{orange}{\circ}\color{magenta}{\bullet}\color{teal}{\bullet}\color{lightgray}{\bullet} \\
\end{array}
\]

The join is of particular importance because it appears when we convert and `opened'
view of a term into its co-de Bruijn counterpart.
%
As we mentioned earlier, co-de Bruijn terms in an arbitrary scope are represented by
the pairing of a term indexed by its precise support with a thinning embedding this
support back into the wider scope.
%
When working with such a representation, it is convenient to have access to an
`opened' view where the outer thinning has been pushed inside therefore exposing
the term's top-level constructor, ready for case-analysis.


The following diagram shows the correspondence between an `opened' application node
using the view (the diamond `\$' node) with two subterms both living in the outer scope
and its co-de Bruijn form (the circular `\$' node) with an outer thinning selecting the
term support.

\noindent
\begin{minipage}{.45\textwidth}\center
  \ExecuteMetaData[ast.tex]{opened}
\end{minipage}\hfill
\begin{minipage}{.45\textwidth}\center
  \ExecuteMetaData[ast.tex]{opening}
\end{minipage}

The outer thinning of the co-de Bruijn term is obtained precisely by
computing the join of the respective outer thinnings of the `opened'
application's function and argument.

These explicit bit manipulations will be preserved during compilation and
thus deliver more efficient code.


\subsection{Compiled Code}\label{sec:compiledview}

The following code block shows the JavaScript code that is produced when compiling the
\IdrisFunction{view} function. We chose to use the JavaScript backend rather than e.g.
the chezscheme one because it produces fairly readable code.
%
We have modified the backend to also write comments reminding the reader of the type
of the function being defined and the data constructors the natural number tags
correspond to.

The only manual modifications we have performed are the inlining of a function
corresponding to a \IdrisKeyword{case} block, renaming variables and property names
to make them human-readable, introducing the \texttt{\$tail} definitions to make
lines shorter, and slightly changing the layout.

\input{view}

Readers can see that the compilation process has erased all of the indices
and the proofs
showing that the invariant tying the efficient runtime representation to the
high-level specification is maintained.
%
A thinning is represented at runtime by a JavaScript object with two properties
corresponding to \IdrisType{Th}'s runtime relevant fields: \IdrisFunction{bigEnd}
and \IdrisFunction{encoding}.
%
Both are storing a JavaScript \texttt{bigInt} (one corresponding to the
\IdrisType{Nat}, the other to the \IdrisType{Integer}).
%
For instance the thinning [01101] would be at runtime
\mintinline{javascript}{{ bigEnd: 5n, encoding: 13n }}.
%

The view proceeds in two steps. First if the \texttt{bigEnd} is \texttt{0n}
then we know the thinning is empty and can immediately return the \IdrisData{Done}
constructor.
%
Otherwise we know the thinning to be non-empty and so we can compute the big end
of its tail (\texttt{\$predBE}) by subtracting one to the non-zero \texttt{bigEnd}.
We can then inspect the bit at position \texttt{0} to decide whether to return a
\IdrisData{Keep} or a \IdrisData{Drop} constructor. This is performed by using a
bit mask to 0-out all the other bits (\texttt{\$th.bigEnd\&1n}) and checking whether
the result is zero.
%
If it is not equal to 0 then we emit \IdrisData{Keep} and compute the \texttt{\$tail}
of the thinning by shifting the original encoding to drop the 0th bit. Otherwise we
emit \IdrisData{Drop} and compute the same tail.

By running \IdrisFunction{view} on this [01101] thinning, we would get
back (\IdrisData{Keep} [0110]), that is to say
\mintinline{javascript}{{ tag: 1, val: { bigEnd: 4n, encoding: 6n } }}.

Thanks to \idris{}'s implementation of Quantitative Type Theory we have managed
to manufacture a high level representation that can be manipulated like a classic
inductive family using smart constructors and views without giving up an inch of
control on its runtime representation.

The remaining issues such as the fact that we form the view's constructors only
to immediately take them apart thus creating needless allocations can be tackled
by reusing Wadler's analysis (section 12 of \cite{DBLP:conf/popl/Wadler87}).
